---
title: "BDP2-448 Lovasz"
date: "`r Sys.Date()`"
author: Benjamin Chan (chanb@ohsu.edu)
output:
  html_document:
    toc: true
    theme: simplex
---

```{r, echo = FALSE, results = "hide", message = FALSE, warning = FALSE}
library("checkpoint")
checkpoint("2019-01-01")
library("magrittr")
library("tidyverse")
library("readxl")
library("broom")
library("knitr")
library("svglite")
library("lme4")
library("nlme")
library("broom.mixed")
library("multcomp")
```

# Import data

```{r}
getwd()
f <- "data/raw/Becky-Bedding Exposure Study Data for Stats.2.xlsx"
```

Import data file *`r f`*

```{r}
df <- f %>% read_excel()
oldnames <- names(df)
newnames <- c("id",
              "sex",
              "time",
              "bedding",
              "order",
              "room",
              "foodIntakeNormalized",
              "orts",
              "foodIntakeRaw",
              "cumulativeFoodIntakeNormalized")
names(df) <- newnames
data.frame(oldnames, newnames) %>% kable()
```

Remove outlier

```{r}
df %>%
  dplyr::select(id, sex, time, bedding, foodIntakeNormalized) %>%
  filter(id == 42 & bedding == "B" & time %in% c(4, 12)) %>%
  kable()
df <-
  df %>%
  filter(!(id == 42 & bedding == "B" & time %in% c(4, 12))) %>%
  mutate(hadOutlierRemoved = case_when(id == 42 & bedding == "B" ~ TRUE,
                                       TRUE ~ FALSE))
```

Plot data.

```{r}
G <-
  df %>%
  mutate(sex = factor(sex,
                      levels = c("F", "M"),
                      labels = c("Females", "Males")),
         bedding = factor(bedding,
                          levels = c("A", "B"),
                          labels = c("Control bedding", "Sick bedding"))) %>%
  ggplot() +
  aes(x = time, y = foodIntakeNormalized, color = bedding) +
  geom_line(aes(group = id, linetype = hadOutlierRemoved), alpha = 1/4) +
  geom_boxplot(aes(group = time)) +
  facet_grid(sex ~ bedding) +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous("Time\n(hours)", breaks = c(2, 4, 12, 24, 36, 48)) +
  scale_y_continuous("Food intake, normalized\n(mg food/g body weight)") +
  theme_bw() +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
ggsave("figures/lineplot.png", dpi = 300)
ggsave("figures/lineplot.svg", dpi = 300)
```

```{r lineplot}
G
```


# New analysis

Prior analysis is archived in commit `f3ce92`.

On Thu, May 2, 2019 at 12:44 PM Becky Lovasz <lovasz@ohsu.edu> wrote:

> Hi Ben,
>  
> My project was put on the back burner for a while, and I'm picking it back up
> again. I discussed the results below with my research advisor. Can you please
> compare normalized food intake for Bedding A vs. Bedding B, regardless of
> bedding exposure order, at each time point separately? We'd like to see males
> only, females only, and all mice combined. Thanks for your help!


```{r}
random <- formula(~ 1 | id)
cs <-
  corSymm(form = random, fixed = FALSE) %>%
  Initialize(data = df)
ctrl <- lmeControl(opt = "optim",
                   maxIter = 500, msMaxIter = 500,
                   tolerance = 1e-6, niterEM = 25, msMaxEval = 200, msTol = 1e-7)
testContrast <- function (nlmeObj, contrast) {
  require(multcomp)
  g <- glht(nlmeObj, linfct = c(contrast)) %>% summary()
  x <- g[["test"]][["coefficients"]]
  dir <- ifelse(x > 0, "higher", "lower")
  s <- g[["test"]][["sigma"]]
  p <- g[["test"]][["pvalues"]][[1]]
  result <- data.frame(contrast = contrast,
                       coefficient = x,
                       sigma = s,
                       pvalue = p,
                       effsize = abs(x),
                       direction = dir,
                       stringsAsFactors = FALSE)
  rownames(result) <- NULL
  result
}
```

## All mice combined

Use `nlme::lme()`.
Specify a general correlation structure with `corSymm()`.
Use this output.

```{r}
fixed <- formula(foodIntakeNormalized ~
                   bedding +
                   time +
                   bedding * time)
M <-
  df %>%
  mutate(time = factor(time)) %>%
  lme(fixed, data = ., random = random, correlation = cs, control = ctrl)
M %>% 
  tidy() %>%
  filter(effect == "fixed") %>%
  kable(digits = 3)
testBeddingEffect <- 
  bind_rows(testContrast(M, "beddingB = 0"),
            testContrast(M, "beddingB + beddingB:time4 = 0"),
            testContrast(M, "beddingB + beddingB:time12 = 0"),
            testContrast(M, "beddingB + beddingB:time24 = 0"),
            testContrast(M, "beddingB + beddingB:time36 = 0"),
            testContrast(M, "beddingB + beddingB:time48 = 0"))
testBeddingEffect %>% kable(digits = 3)
test1 <- testBeddingEffect
```

## By sex

Use `nlme::lme()`.
Specify a general correlation structure with `corSymm()`.
Use this output.

```{r}
fixed <- formula(foodIntakeNormalized ~
                   bedding +
                   time +
                   sex +
                   bedding * time +
                   bedding * sex +
                   sex * time +
                   bedding * time * sex)
M <-
  df %>%
  mutate(time = factor(time)) %>%
  lme(fixed, data = ., random = random, correlation = cs, control = ctrl)
M %>% 
  tidy() %>%
  filter(effect == "fixed") %>%
  kable(digits = 3)
testBeddingEffect <- 
  bind_rows(testContrast(M, "beddingB = 0"),
            testContrast(M, "beddingB + beddingB:time4 = 0"),
            testContrast(M, "beddingB + beddingB:time12 = 0"),
            testContrast(M, "beddingB + beddingB:time24 = 0"),
            testContrast(M, "beddingB + beddingB:time36 = 0"),
            testContrast(M, "beddingB + beddingB:time48 = 0"),
            testContrast(M, "beddingB + beddingB:sexM = 0"),
            testContrast(M, "beddingB + beddingB:sexM + beddingB:time4:sexM = 0"),
            testContrast(M, "beddingB + beddingB:sexM + beddingB:time12:sexM = 0"),
            testContrast(M, "beddingB + beddingB:sexM + beddingB:time24:sexM = 0"),
            testContrast(M, "beddingB + beddingB:sexM + beddingB:time36:sexM = 0"),
            testContrast(M, "beddingB + beddingB:sexM + beddingB:time48:sexM = 0"))
testBeddingEffect %>% kable(digits = 3)
test2 <- testBeddingEffect
```


# Interpretation

Using a model that does not distinguish male and female mice, there is a
difference in food intake by bedding type at 4 and 24 hours.
At 4 hours, food intake is
`r test1 %>% filter(contrast == "beddingB + beddingB:time4 = 0") %>% pull(effsize) %>% sprintf("%.3g", .)` mg/g
`r test1 %>% filter(contrast == "beddingB + beddingB:time4 = 0") %>% pull(direction)`
with bedding type B compared to A
(p-value: `r test1 %>% filter(contrast == "beddingB + beddingB:time4 = 0") %>% pull(pvalue) %>% sprintf("%.2g", .)`).
At 24 hours, food intake is
`r test1 %>% filter(contrast == "beddingB + beddingB:time24 = 0") %>% pull(effsize) %>% sprintf("%.3g", .)` mg/g
`r test1 %>% filter(contrast == "beddingB + beddingB:time24 = 0") %>% pull(direction)`
with bedding type B compared to A
(p-value: `r test1 %>% filter(contrast == "beddingB + beddingB:time24 = 0") %>% pull(pvalue) %>% sprintf("%.2g", .)`).

Among female mice, there is a difference in food intake by bedding type at 24
and 48 hours.
At 24 hours, food intake is
`r test2 %>% filter(contrast == "beddingB + beddingB:time24 = 0") %>% pull(effsize) %>% sprintf("%.3g", .)` mg/g
`r test2 %>% filter(contrast == "beddingB + beddingB:time24 = 0") %>% pull(direction)`
with bedding type B compared to A
(p-value: `r test2 %>% filter(contrast == "beddingB + beddingB:time24 = 0") %>% pull(pvalue) %>% sprintf("%.2g", .)`).
At 48 hours, food intake is
`r test2 %>% filter(contrast == "beddingB + beddingB:time48 = 0") %>% pull(effsize) %>% sprintf("%.3g", .)` mg/g
`r test2 %>% filter(contrast == "beddingB + beddingB:time48 = 0") %>% pull(direction)`
with bedding type B compared to A
(p-value: `r test2 %>% filter(contrast == "beddingB + beddingB:time48 = 0") %>% pull(pvalue) %>% sprintf("%.2g", .)`).

Among male mice, there is a difference in food intake by bedding type at 48 hours.
At 48 hours, food intake is
`r test2 %>% filter(contrast == "beddingB + beddingB:sexM + beddingB:time48:sexM = 0") %>% pull(effsize) %>% sprintf("%.3g", .)` mg/g
`r test2 %>% filter(contrast == "beddingB + beddingB:sexM + beddingB:time48:sexM = 0") %>% pull(direction)`
with bedding type B compared to A
(p-value: `r test2 %>% filter(contrast == "beddingB + beddingB:sexM + beddingB:time48:sexM = 0") %>% pull(pvalue) %>% sprintf("%.2g", .)`).


# R session information

For debugging purposes.

```{r, echo = FALSE}
sprintf("Run time: %s", Sys.time())
print(sessionInfo(), locale = FALSE)
```
